---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CNpare

## Summary of the package

**CNpare** is designed to compare and contrast genome-wide tumour copy number profiles. By using CNpare, one can:

-   Calculate similarities between copy-number profiles
-   Quantify the percentage of genome differences between copy-number profiles
-   Compare copy-number profiles based on copy-number signatures

## Installation

You can install the development version of CNpare from this github repository using devtools package. Once installed, CNpare can be loaded:

```{r, eval=FALSE}
library(CNpare)
```

## Workflow overview

### Input data

The most basic initial input to the CNpare package consists of a data frame containing the copy number profile segmented from a tumour sample.

This structure must contain the following columns:

-   chromosome -- `chromosome`
-   start position -- `start`
-   end position -- `end`
-   copy number -- `segVal`
-   sample identifier -- `sample`

Alternatively, a `QDNAseqCopyNumbers` object can be also used as input.

The segmented profile used as input needs to be transformed to bin-level copy number profiles. This can be done using `getCNbins()`. Genomic positions of bins can be obtained using `getBinsStartsEnds()`.

This package includes `cells_segcn.RData`, a dataframe with copy-number profiles of 1,417 human cancer cell lines from the Cancer Cell Line Encyclopaedia (CCLE) and Genomics of Drug Sensitivity in cancer (GDSC), which has been generated with ASCAT. This dataset must also be transformed to bin-level copy numbers before to be used for comparisons.

### Compare two copy-number profiles

The function `getSimilarities()` can be used to get all similarity metrics calculated by CNpare. This function needs at least two bin-level copy number profiles to compare.

Four similarity metrics are computed for comparison by default, but the user can be interested in only computing one of them. It can be done by modifying the `method` parameter as follows:

-   Pearson correlation -- `method=pearson`
-   Manhattan distance -- `method=manhattan`
-   Euclidean distance -- `method=euclidean`
-   Cosine similarity -- `method=cosine`

### Compute the % genome differences

Using the function `getDifference()`, the extent of genome differences (%) between two profiles can be calculated

### Cluster profiles based on their exposure to copy-number signature

First, CNpare extracts the activity levels of the 7 ovarian copy-number signatures using the computational approach developed by *Macintyre el al. Nature Genetics, 2018*. Functions used for this purpose are:

-   `extractCopynumberFeatures()`
-   `generateSampleByComponentMatrix()`
-   `quantifySignatures()`

Data needed for quantifying these signatures is included in CNpare, except for `component_parameters.rds` which may be downloaded from here: https://bitbucket.org/britroc/cnsignatures/src/master/data/. We are preparing an experiment data package for including all data needed for identifying copy-number signatures

Finally, using the function `plotClusters()`, profiles are clustered according to their exposure to each copy number signature.

## Manuscript analysis Rmarkdown

The two `Rmarkdown` documents can be compiled to reproduce the entire analysis carried out in the manuscript:

-   `CNpare_AssessingTool` -- this tutorial reproduces the analyses performed to check the ability of the tool to find the most similar profile
-   `CNpare_DemonstratingUtility` -- this tutorial follows the typical workflow of the tool

To compile this document run knitr to html. Packages required to compile the Rmarkdown document include: knitr, dplyr, ggplot2, reshape2, data.table, factoextra, flexmix, NMF, stats, splitstackshape, kableExtra and magrittr.
